---
title: "Week 7 Assignment"
author: "Ted Laderas"
date: "8/6/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)

analytic_table <- readr::read_csv("analytic.csv", 
                                    col_types = cols(
                                      encounter_type = col_factor(NULL),
                                      outcome = col_factor(NULL),
                                      Admit_source= col_factor(NULL),
                                      indexadmit = col_factor(NULL),
                                      Readmit30 = col_factor(NULL)
                                    ))
```

In this assignment, we're going to focus on visualizing the readmissions and outcomes.

## Problem 1

Show the `outcome` by month of `Admit_date` as a dodged bar plot for `analytic_table`. Save the result as `outcome_graph`. (Hint: think about what needs to be mapped to `x` and `fill`.) Make sure to add a title.

```{r}
#space for your answer here
```

Do you notice anything special?

## Problem 2

As a bar graph, show the `Admit_source` on the `x` axis and `Readmit30` as a fill.

```{r}
#space for your answer here
```

## Problem 3

Produce a histogram of `length_of_stay` and color by `Readmit30`. 

```{r}

```

## Problem 4

Use `patchwork` to arrange your graphs. You might use all 3, or only two. Do we know anything about readmits? What do we know? (answer in text below your plot)

```{r}

```

text answer here.