---
title: "Prework Assignment"
author: "David Dorr/Tracy Edinger/Ted Laderas"
date: "June 28, 2016"
output: pdf_document
---

**Purpose:**  To build and test your data warehouse from extracted files.

**Background:**  In the analytics class, you will have a miniature data warehouse that will be accessed by R and SQLite.  In order to do this, you will load a set of 7 flat files into a SQLite database, which will serve as your data warehouse.  After you do this, you will need to record key information in your data dictionary about these tables.

Load the flat files `patient_encounter_hosp.txt` and `patient.txt` into a SQLite database using R; these files are both delimited by a pipe “|” and contain variable names in the header.  Use a SQL query to return the number of readmissions within 30 days contained in the dataset; note that there is not a variable for readmissions, but the file is limited to initial hospitalizations and readmissions.  Use an R query or SQL to summarize the readmission rate, either by count or percent or both. 

Some hints:  The `patient_encounter_hosp` file should have one to two observations per patient, each observation a specific hospitalization.  The analyst has subset the file for you to assure there are only INDEX admissions and READMISSIONS.  They do not look different, however, so you'll have to do some date differences across patientid to find readmissions.

Eventually, you will receive a data dictionary which you will improve; in the interest of fairness, here it is now.  I think it is not useful for this assignment, but you may not agree.

###Deliverables:   

1) R Code that loads the data warehouse; and 
2) R Code that encodes index admissions with whether a readmission follows (and discarding readmissions).
3) Data Dictionary for each table (excel file is fine).

Source : All 7 flat files provided here. 

##Suggested Steps:

1) Download the 7 files ([zip file](https://ohsu.box.com/shared/static/xqios7ryxzfj7ztgc1y77hz68c94gni9.zip)) if you haven't already.

2) Write code that loads all the flat files (all with header rows and with a pipe delimiter).  

3) Store all the data frames into a database in R and SQLite.  Note that you’ll need to load the RSQLite or the sqldf package from a Repository and install it first.  My example uses sqldf, which I found simpler.  

4) Create a new table in your SQLite database by using querying SQL to identify just the index admissions (the first admission dates) and call this variable `indexadmit`. *Hint*: which table should you use for this? How can you use a self-join to identify the earliest dates for each patient?

5) Next, identify those admission dates that are readmissions within 30 days from the index admission. Name the readmission variable `Readmit30`.  Note that the file only has the index admission and readmission within 30 days; it does so by limiting rows for individual patients to maximum 2 admissions (either a single admission or the admission and the readmission within 30 days).

5) Create a new data dictionary page for your analytic file, and record your first variables in the data dictionary format See example data dictionary 

6) Report the Count of unique patients and the number of readmissions in the file.

# Hints:  

1) You might create a new directory and use the R command `setwd()` to make this easier.
2) the `read.table()` function will help here, and to 'assign' the table to a data.frame use the syntax of 

```{r}
dataframename <- read.table(filename)  
```

http://www.inside-r.org/r-doc/utils/read.table  

3) You've done a lot of this for R, but you probably want to save a data set in R;  SQLite requires several steps to assign a database properly.  You might check a few resources, starting with http://cran.r-project.org/web/packages/RSQLite/RSQLite.pdf  but don't end there.  You'll need some code like

```{r}
drv <- dbDriver("SQLite")
tfile <- tempfile()
con <- dbConnect(drv, dbname = tfile)
dbWriteTable(con, "patient", patient)
```

4) You'll need to embed your SQL query into a command for SQLite with something like dbGetQuery.  Note that SQLite does not have a date data type.  Ha.  This requires either pre-processing dates in R for your best schema or taking substr of the dates to format it in the date functions within SQLite.

5) You have a series of choices to make here.  Improve the schema you received. Still stuck?  Try this help file Task 1.a for your R script 

