---
title: 'Task 3: SQLite Introduction'
author: "Ted Laderas"
date: "June 9, 2016"
output: pdf_document
---

This week we will be loading our data into a database type called SQLite. SQLite is a database engine that's open-source and is widely used. We'll use it to simulate our data warehouse that we will need to calculate our analytics on. Loading into a database system will also let us produce more complicated queries that join the different tables in our dataset. 

##Installing the SQLite Package

The first thing you need to do is install the RSQLite package into R. This package contains the SQLite engine and will let you interact with SQLite databases directly.

```{r eval=FALSE}
install.packages("RSQLite")
```

Now that you have the RSQLite package, you need to load it up with the `library()` command.

```{r}
library(RSQLite)
```

##Initializing the Database File for Our Data Warehouse

The first thing we need to do is initialize our database file and connect to it. SQLite store an entire database into a single file. By using the `dbConnect()` command, we initialize a new SQLite database called "patient.sqlite". Note that if this database file already existed in our working directory, then we'd just be connecting to it.

```{r}
SQLiteConnection <- dbConnect(drv=SQLite(),dbname="patient.sqlite")
```

Let's load in one of our tables into R first and then save it as a table into the database:

```{r}
patient <- read.table("RawSyntheticData/patient.txt", header=TRUE, sep="|")

dbWriteTable(conn = SQLiteConnection, "patient", patient)

#in order to save our table, we must disconnect from the database
dbDisconnect(SQLiteConnection)
```

Confirm that you saved the data in your SQLite database correctly by running a query on the table. Here we're just counting the number of patients by counting the number of patient IDs.

The first thing we do is set up the query, which counts the number of rows in our newly saved table in our database.

```{r}
#show number of rows in table
nrow(patient)

#initialize database connection again
SQLiteConnection <- dbConnect(drv=SQLite(),dbname="patient.sqlite")

#Here we specify our query 
sqlStatement <- "select count(*) from patient"

#confirm the answer for the database is the same as the table in R
res <- dbGetQuery(SQLiteConnection, sqlStatement)

#show result
res

#note that if our result returns a table, we need to use the fetch() command to
#return it as a data.frame (more about this next week).

#Confirm the two are equal
nrow(patient) == res
```

##This Week's Assignment

Load the rest of your tables into your SQLite Database. Construct a query for each table to confirm that the number of rows of each loaded table is identical to the number of rows in the corresponding table you loaded into R.  

Submit a copy of your data loading script, with comments.