---
title: 'Task 4: SQLite: Joining Tables and the CASE statement'
author: "Ted Laderas"
date: "June 9, 2016"
output: pdf_document
---

This week we will be querying our database. We'll also be doing simple joins and queries to show the power of SQLite. We'll also be using the CASE statement to identify cases that meet our criteria.

If you haven't done last week's assignment (loading the dataset into SQLite), please do it now.

## SQL Queries with Joins

Now that we have our SQLite database (our "data warehouse"), we can now start asking more complicated queries. 

For example: *How many high risk patients had emergency room admissions?*

In order to do this, we will do what is called a database join, where we map rows in one table (in this case the `patient` table) to another table (`patient_encounter_hosp`) through a mapping key, `patientid`. After we do the join, we can use our filtering criteria as before. We do this by joining our criteria using a "where" statement.

Here are the first few rows of the `patient` table to remind you of what's in it. What variable are we looking for in this table?

```{r echo=FALSE}
tab <- read.table("RawSyntheticData/patient.txt", sep="|", header = TRUE)
head(tab)
```

Here are the first few rows of the `patient_encounter_hosp` table to remind of what's in it. What variable are we looking for in this table?

```{r echo=FALSE}
tab <- read.table("RawSyntheticData/patient_encounter_hosp.txt", sep="|", header=TRUE)
head(tab)
```

Now that you have seen both tables, we will need to combine them into a single table to query using a join. The common element between the two tables that we will join on is `patientid`, or the patient identifier. Note that this variable name is the same for both tables. In order to specify the join, we will need to specify the origin of both variables by using the `.` operator. That is, if we want to use the `patientid` column from the `patient` table, we need to specify it as `patient.patientid`.

```{r warning=FALSE}
#load the RSQLite library
library(RSQLite)

#connect to our database
#if you have named your database differently, change the dbname argument
SQLiteConnection <- dbConnect(drv=SQLite(),dbname="patient.sqlite")

#merge the two tables on the common id
#note that we have to specify the origins of each variable to merge
#with the `.` operator. For readablity, it's always good practice
#to specify the origin of a variable name.

sqlStatement <- "SELECT patient.patientid, riskCat, Admit_source, outcome 
                 FROM patient, patient_encounter_hosp
                 WHERE patient.patientid = patient_encounter_hosp.patientid
                 AND patient.riskCat = 'High' 
                 AND patient_encounter_hosp.Admit_source = 'Emergency Room'"

queryResult <- dbGetQuery(SQLiteConnection, sqlStatement)

#show first ten rows of result
queryResult[1:10,]

#What were the outcomes for these patients?
table(queryResult$outcome)
```

One good shortcut that can prevent you from typing too much in your queries is creating *aliases* (or shortcuts) for a table name using the `AS` statement. Here we rewrite the above statement by specifying the alias `p1` for the `patient` table and the alias `peh` for the `patient_encounter_table`. Note that even though the `SELECT` statement goes before our declaring the aliases, it's ok to use the the alias anyway.

```{r}
sqlStatement <- "SELECT p1.patientid, riskCat, Admit_source, outcome 
                 FROM patient AS p1, patient_encounter_hosp AS peh
                 WHERE p1.patientid = peh.patientid
                 AND p1.riskCat = 'High' 
                 AND peh.Admit_source = 'Emergency Room'"

queryResult <- dbGetQuery(SQLiteConnection, sqlStatement)

#show first ten rows of result
queryResult[1:10,]
```

Strictly speaking, we don't need `AS` to declare an alias, we can just put the alias following the table or variable of interest. The following is identical to the above query, except we provide a new alias for `Admit_source` (`ads`) and `outcome` (`o`):

```{r}
sqlStatement <- "SELECT p1.patientid, riskCat, Admit_source ads, outcome o 
                 FROM patient p1, patient_encounter_hosp peh
                 WHERE p1.patientid = peh.patientid
                 AND p1.riskCat = 'High' 
                 AND peh.Admit_source = 'Emergency Room'"

queryResult <- dbGetQuery(SQLiteConnection, sqlStatement)

#show first ten rows of result
queryResult[1:10,]
```

**QUESTION**: How would we modify this query to select `Normal` Risk patients who had an outcome of `Rehab`?

## The CASE statement

Sometimes you just want to code whether a row in a table meets a particular criteria or not. This can be useful when chaining queries or as a 'dummy' variable in statistical analysis. You can do this by using the `CASE` statement. 

The case statement has the following syntax:

```{eval=FALSE}
CASE WHEN [condition] 
     THEN [expression1] 
     ELSE [expression2] END 
```

In the `[condition]` slot, we can add any filtering criteria. If our `[condition]` is met, then the SQL statement executes `[expression1]`. If it isn't met, then the SQL statement executes `[expression2]`. 

Let's modify our above statement to use a case statement instead of a WHERE statement, such that our new variable will equal 1 (which is our `[expression1]`) when it equals the query (`patient.riskCat = 'High' AND patient_encounter_hosp.Admit_source = 'Emergency Room'`) and 0 otherwise (which is our `[expression2]`). We define a new variable `highRiskER` in our returned results by specifying it after the CASE statement.

```{r}
#note the comma before the case statement!
#we are 
sqlStatement <- "SELECT patient.patientid, riskCat, Admit_source, outcome, 
                 CASE 
                    WHEN
                      patient.riskCat = 'High'  
                      AND patient_encounter_hosp.Admit_source = 'Emergency Room'
                    THEN 1
                 ELSE 0 
                 END highRiskER
                 FROM patient, patient_encounter_hosp
                 WHERE patient.patientid = patient_encounter_hosp.patientid"
                 

queryResult <- dbGetQuery(SQLiteConnection, sqlStatement)

#confirm that our case statement worked (there are two rows here that meet the criteria)
queryResult[20:50,]

#another way to confirm the query worked 
#confirm that those cases with highRiskER == 1 meet our criteria
filteredResult <- queryResult[queryResult$highRiskER == 1,]

filteredResult[1:20,]
```

Compare this statement to the previous SQL statement. Note the placement of the `CASE` statement compared to the placement of the the `WHERE` statement in the two queries.

Now that we have our `highRiskER` coded, we can ask some questions that require crosstables. For example, how many high risk patients have an `SNF` outcome (i.e., ended up in a Skilled Nursing Facility) versus normal patients? Here I use the `table()` function built into R, but you can also write a SQL query to do this using CASE and COUNT statements. In order to do so, I simply build a crosstable between the `highRiskER` and the `outcome` table.

```{r}
#note that we have to specify the variable names using the $ operator
table(queryResult$highRiskER, queryResult$outcome)
```

**QUESTION:** How do we modify the above query for the a case variable `NormalClinic` that returns a 1 where the riskCategory is `Normal` and the admit source is `Clinic`?

**QUESTION:** Write a query that selects those High Risk patients who have a diagnosis of Acute myocardial infarction of anterolateral wall (ICD9 code is 410.00). How many are there?

## Further Resources

For more information, please refer to [Module 5](https://github.com/laderast/r-bootcamp/tree/master/module5) of the R-Bootcamp.
