---
title: 'Task 2: Loading External Datasets/Descriptive Statistics'
author: "Ted Laderas"
date: "June 9, 2016"
output: pdf_document
---

## Goals for this week

This week we will download the datasets, load them into R, and do some more simple descriptive statistics on them.

Download the [synthetic dataset here](https://sakai.ohsu.edu/access/content/group/BMI-569-1-AF-Su17/modules/RawSyntheticData.zip), unzip it, and place the folder somewhere easily accessible (the Desktop is fine for now). 

You will want to start an RStudio project in this folder, since you will be accessing this data for the rest of the pre-work and the in-class sessions. (Go to `>File>New Project>Create Project in Existing Directory` and select the folder.) 

Once you do this, you can just click on the `.Rproj` file it generates and have everything easily accessible within the folder. R will able to "see" all of the data files in this folder.

## The concept of a working directory

Let's load up Rstudio again and your project. You can find out what your *working directory* is with the `getwd()` command.

```{r}
getwd()

#list all the files in my current working directory 
#(these are the files in my current working directory)
list.files()
```

You can set your working directory using the `setwd()` command. Here my data files are in the `~/Code/AnalyticsCourse/RawSyntheticData` folder, so I'll set it there:

```{r}
setwd("~/Code/AnalyticsCourse/RawSyntheticData/")
#confirm R can 'see' your files
list.files()

#change it back
setwd("~/Code/AnalyticsCourse/")
```

## Loading in Data

Now that we've set our working directory, we can now load our data. Before you do this, you will want to check the readme file to let you know if there are any issues.

Opening the file `patient.txt` in a text editor, we notice a few things. The fields are separated by the `|` character, so we need to specify that. Also, there is a header for the file.

```{r}
patient <- read.table("RawSyntheticData/patient.txt", sep = "|", header=TRUE)

#confirm that we have loaded the data directly.
head(patient)
```

Now that we have the data loaded, we can confirm some properties of the data. Let's do a `summary()` first to get a basic idea of what the data looks like. How many patients are High risk versus Normal Risk? (Hint: look at the `riskCat` field.)

```{r}
summary(patient)
nrow(patient)
colnames(patient)
```

In order to access the individual columns, we can use the `$` operator to do operation on the individual columns. Here I'm running a simple `table()` operation to count the number of rows within each category for the `riskCat` variable.

```{r}
table(patient$riskCat)
```

## This Week's Assignment (1 point)

Load in the `patient_encounter.txt` file into R, and confirm that you loaded it correctly. Submit your answers as a .R file, and show the code you used and your
answers as comments (use `##` to show which lines are your answers)

1) How many `No shows` were there in the dataset? (Look in the outcome field). 
2) If we want to count the number of readmissions, would we want patients with these values?

## How to submit

Submit your answer using the `week2Submission.Rmd` file. This is an RMarkdown file and you can mix code and text using it. You can open this file directly into Rstudio. 

Make sure to rename the `week2Submission.Rmd` file to `week2Submission-LASTNAME.Rmd` with your last name, so we can confirm that you submitted the assignment.

## For More Information

Please refer to R-Bootcamp, [module 2](https://github.com/laderast/r-bootcamp/tree/master/module2) for more information about `data.frames`.
