---
title: "Week 5-6 Assignment"
author: "Put your name here"
date: "6/21/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

## Query Code

Your final problem in the prework is to identify the `index_admit` cases in the `patient_encounter_table` which have a readmission within 30 days of the `index_admit`'s discharge date. Code this as a case statement and save the results as another variable called `readmit30` in your table. 

There are multiple routes to do this: You can first identify the index cases, save that version of `patient_encounter_hosp` as a table in your database, and then identify the cases within the index variable. However, you can also combine the two cases into a single statement, because we know that there are only two types of entries in these table. How could you do that? (Hint: you will have to do another self-join.)

Put your query code below.

```{r}
##show your code here

```

## Question 1

1) How many index cases are there?

Answer: 

```{r}
## put your code here
```

## Question 2

2) How many readmit30 cases are there?

Answer: 

```{r}
## put your code here
```
