---
title: "Assignment 1 Answer Key"
author: "Answer Key"
date: "6/21/2019"
output: html_document
---

```{r setup, include=FALSE}
#ignore this section, it's just there for formatting.
knitr::opts_chunk$set(echo = TRUE)

```

## Assignment Week 1 (1 point)

Answer these questions about the `patient` table in the database. Show your code in your submitted notebook that you used to answer the question.

Note that I am using an extra package called `here`. This has to do with how notebooks look for files. They always look for files *relative* to their location. `here` lets you always specify the working directory as the root directory of your project. The way I've set up your assignments, you won't need to use them.

1) How many rows are there?

```{r}
library(RSQLite)
library(tidyverse)
library(here)
#start our connection to the database
con <- DBI::dbConnect(RSQLite::SQLite(), here("data/patient1.sqlite"))
patient_table <- tbl(con, "patient") 
patient_table_inmemory <- collect(patient_table)
##put the code that you used to find the answer to question 1 here.

```

## Question 2

What variables are missing all information (represented by NAs) and why might it be missing?

```{r}
##put the code that you used to find the answer to question 2 here.

summary(patient_table_inmemory)
```

## Question 3

3) What is the maximum `riskscore` value?

```{r}
##put the code that you used to find the answer to question 3 here.
##Don't forget to disconnect!
dbDisconnect(con)
```

## If you need help

Please post questions to the BMI569 slack, or email Ted Laderas or Aaron through sakai.