---
title: "Week 2 Assignment"
author: "Put your name here"
date: "6/21/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Question 1

For the `patient_encounter` table, produce a data dictionary (you can use the format above in the `week2submission.nb.html`, or save it as an excel file). Submit your data dictionary as part of the assignment.

```{r}
##show the code you used to find this answer here
library(RSQLite)
library(here)
library(tidyverse)
#start our connection to the database
con <- DBI::dbConnect(RSQLite::SQLite(), here("data/patient1.sqlite"))

patient_encounter <- tbl(src=con, from="patient_encounter")
patient_encounter <- collect(patient_encounter)
dbDisconnect(con)

```

Run `summary()` - are there any `character` or `numeric` variables that might be categories? Recast them as `factors` by using `mutate`.

```{r}

```

Produce a visual summary (visdat) and a skimr summary of the data. Are there any things you notice about this table?

```{r}

```

Add your data dictionary as text here, or submit as an excel file.


## Question 2

How many `No shows` were there in the dataset? (Look in the `outcome` variable). 

Answer:

```{r}
##show the code you used to find this answer here


```

## Question 3

We will want to predict 30 day readmissions in our dataset, which means that we want to predict whether a patient will be readmitted to the hospital within 30 days of being readmitted. 

There are multiple reasons for a patient being readmission, but we're interested in the ones that we can prevent. If we want to count the number of readmissions, would we want to include patients who are just `No shows`?

Answer (no need to show code here):




## Question 4

Construct a query for each table in the database to output the number of rows for each table. 

```{r}
##show your code here

```

## Question 5

How many patient encounters in in `patient_encounter_hosp` were from the emergency room and were discharged home? 

```{r}
##show your code here
```

## Question 6

How many patients in `patient_encounter` had complete outcomes and an encounter type of 48?

```{r}
##show your code here
```

## Question 7

Based on your exploration and data dictionary of the `patient_encounter` table last week, come up with another query that uses the AND statement to chain two other criteria (it is probably easiest to use two categorical variables).

```{r}
##show your code here
```


## How to submit

Submit your html from knitting the `week2Submission.Rmd` file. This is an RMarkdown file and you can mix code and text using it. 

Make sure to rename the `week2Submission.html` file to `week2Submission-LASTNAME.html` with your last name, so we can confirm that you submitted the assignment.
