---
title: 'Week 2: Loading External Datasets/Descriptive Statistics'
author: "Ted Laderas"
date: "June 9, 2016"
output: pdf_document
---

## Goals for this week

This week we will download the datasets, load them into R, and do some more simple descriptive statistics on them.

Download the [analyticsWorkspace zip](https://sakai.ohsu.edu/access/content/group/BMI-569-1-AF-Su17/modules/RawSyntheticData.zip), unzip it, and place the folder somewhere easily accessible (the Desktop is fine for now). 

Open the `analyticsWorkspace.Rproj` file. It should open this project up in RStudio. This project is where you will be doing the rest of your work for the Analytics Course.

## The concept of a working directory

Let's load up Rstudio again and your project. You can find out what your *working directory* is with the `getwd()` command.

```{r}
getwd()

#list all the files in my current working directory 
#(these are the files in my current working directory)
list.files()
```

## Loading in Data

Now that we've set our working directory, we can now load our data. Before you do this, you will want to check the readme file to let you know if there are any issues.

Opening the file `patient.txt` in a text editor, we notice a few things. The fields are separated by the `|` character, so we need to specify that. Also, there is a header for the file.

```{r}
patient <- read.table("data/patient.txt", sep = "|", header=TRUE)

#confirm that we have loaded the data directly.
head(patient)
```

Now that we have the data loaded, we can confirm some properties of the data. Let's do a `summary()` first to get a basic idea of what the data looks like. How many patients are High risk versus Normal Risk? (Hint: look at the `riskCat` field.)

```{r}
summary(patient)
nrow(patient)
colnames(patient)
```

In order to access the individual columns, we can use the `$` operator to do operation on the individual columns. Here I'm running a simple `table()` operation to count the number of rows within each category for the `riskCat` variable.

```{r}
table(patient$riskCat)
```

## Knitting Your Submission File

One of the great things about using `.Rmd` (Rmarkdown) files is that you can make reports that you can rerun. We will use a package called `rmarkdown` to do this.

```{r eval=FALSE}
##install knitr if you haven't already
install.packages("rmarkdown")
```

Once `rmarkdown` is installed, try running the following commands in the console:

```{r eval=FALSE}
library(rmarkdown)
rmarkdown::render("week2Submission.Rmd")
#you can also use the "Knit" button above this script (it has a ball of yarn next to it).
```

Now open the `week2Submission.html` file that it generated. This is known as `Rmarkdown` output. You can see that it renders both `R` code and text into a single report. It's pretty slick and will prevent you from having to completely put together reports from scratch if the underlying data changes.

## Notes on Rmarkdown

Note that your code must work from scratch (that is, your code must load data and execute code before you can render the markdown file). When `render()` works, it starts from a completely blank workspace and it doesn't see anything in your current workspace. 
I've started you out with the data loading in your `week2Submission.Rmd` script so you don't have to worry about it this time.

The reason for this is to encourage *reproducibility* - having code that executes from raw data to final output is desirable. Note that we usually will have code in multiple stages, so we can use intermediate outputs, such as the database we'll be creating in Week 3.

More info on RMarkdown is here: http://rmarkdown.rstudio.com

## This Week's Assignment (2 points)

Load in the `patient_encounter.txt` file into R, and confirm that you loaded it correctly. Submit your answers as a .R file, and show the code you used and your
answers as comments (use `##` to show which lines are your answers)

1) How many `No shows` were there in the dataset? (Look in the outcome field). 
2) If we want to count the number of readmissions, would we want patients with these values?

## How to submit

Submit your answer using the `week2Submission.Rmd` file. This is an RMarkdown file and you can mix code and text using it. You can open this file directly into Rstudio. 

Make sure to rename the `week2Submission.Rmd` file to `week2Submission-LASTNAME.Rmd` with your last name, so we can confirm that you submitted the assignment.

## For More Information

Please refer to R-Bootcamp, [module 2](https://github.com/laderast/r-bootcamp/tree/master/module2) for more information about `data.frames`.
