---
title: "Week 4 Assignment"
author: "Put your name here"
date: "6/21/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Question 1

Modify a query to create a case variable `NormalClinic` that returns a 1 where the riskCategory is `Normal` and the admit source is `Clinic` (we give you the base query in `week4Submission.Rmd`. Make sure that your query runs on your SQLite Database.

```{r}
##show your code here

sqlStatement <- "SELECT patient.patientid, riskCat, Admit_source, outcome, 
                 CASE 
                    WHEN
                      patient.riskCat = 'High'  
                      AND patient_encounter_hosp.Admit_source = 'Emergency Room'
                    THEN 1
                 ELSE 0 
                 END highRiskER
                 FROM patient, patient_encounter_hosp
                 WHERE patient.patientid = patient_encounter_hosp.patientid"
```


## Question 2

2) Write a query that selects those High Risk patients who have a diagnosis of Acute myocardial infarction of anterolateral wall (ICD9 code is 410.00). How many are there?

Answer: 

```{r}
## put your code here
```
