---
title: "Week 4-5 Assignment"
author: "Put your name here"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

## Query Code

Your final problem in the prework is to identify the `index_admit` cases in the `patient_encounter_table` which have a readmission within 30 days of the `index_admit`'s discharge date. Code this as a case statement and save the results as another variable called `readmit30` in your table. 

Your final table should look like this:

```{r echo=FALSE}
tab <- read.delim("peh_example.txt")
tab
```

There are multiple routes to do this: 

1) You can first identify the index cases, save that version of `patient_encounter_hosp` as a table in your database, and then identify the cases within the index variable. 

2) OR, you can also combine the two cases into a single statement, because we know that there are only two types of entries in these table. How could you do that? (Hint: you will have to do another self-join.)

Put your query code below. This query should be able to run and you should show the first few rows of your query.

```{r}
##show your code here


## When you're done, save your query to the database as "analytic_table" and also save it to disk as `analytic_table.csv`.

## Don't forget to disconnect!
```

## Be Patient

This is a challenging query to get right, which is why we give you two weeks. If you need more help putting this together, consult the `hintsWeek45.pdf` file in this assignment.

Don't hestitate to reach out either by Slack or come to the drop-in sessions. 

## Question 1

1) How many index cases are there?

Answer: 

```{r}
## put your code here
```

## Question 2

2) How many readmit30 cases are there?

Answer: 

```{r}
## put your code here
```
