---
title: 'Task 3: SQLite Introduction'
author: "Ted Laderas"
date: "June 9, 2016"
output: pdf_document
---

```{r echo=FALSE, warning=FALSE, message=FALSE}
if(file.exists("patient1.sqlite")){
  file.remove("patient1.sqlite")
}
```

This week we will be loading our data into a database type called SQLite. SQLite is a database engine that's open-source and is widely used. We'll use it to simulate our data warehouse that we will calculate our analytics on. Loading into a database system will also let us produce more complicated queries that join the different tables in our dataset. 

If you have not done last week's assignment, please do it now. You need to at least have downloaded the synthetic data set and made an RStudio project for that folder.

## Installing the SQLite Package

The first thing you need to do is install the RSQLite package into R. This package contains the SQLite engine and will let you interact with SQLite databases directly.

```{r eval=FALSE}
install.packages("RSQLite")
```

Now that you have the RSQLite package, you need to load it up with the `library()` command. Note that you will have to do this each time you want to use the package.

```{r warning=FALSE}
library(RSQLite)
```

## Initializing the Database File for Our Data Warehouse

The first thing we need to do is initialize our database file and connect to it. SQLite store an entire database into a single file. By using the `dbConnect()` command, we initialize a new SQLite database called `patient1.sqlite`. Note that if this database file already existed in our working directory, then we'd just be connecting to it, rather than creating it.

```{r}
SQLiteConnection <- dbConnect(drv=SQLite(),dbname="patient1.sqlite")
```

Let's load in one of our tables into R first and then save it as a table into the database:

```{r}
patient <- read.table("RawSyntheticData/patient.txt", header=TRUE, sep="|")

#write a table named "patient" with our patient table into the database
dbWriteTable(conn = SQLiteConnection, name="patient", value=patient)

#in order to save our table, we must disconnect from the database
dbDisconnect(SQLiteConnection)
```

Let's confirm that you saved the data in your SQLite database correctly by running a query on the table. Here we're just counting the number of patients by counting the number of patient IDs.

The first thing we do is set up the query, which counts the number of rows in our newly saved table in our database. By using `dbGetQuery()`, we run the query and fetch the result as a `data.frame`. 

```{r}
#show number of rows in table
nrow(patient)

#initialize database connection again
SQLiteConnection <- dbConnect(drv=SQLite(),dbname="patient1.sqlite")

#Here we specify our query 
sqlStatement <- "select count(*) from patient"

#confirm the answer for the database is the same as the table in R
res <- dbGetQuery(SQLiteConnection, sqlStatement)

#show result
res

#Confirm the two are equal
nrow(patient) == res
```

Note that if you think the result of your query will return a large data table (in the millions of rows), using `dbGetQuery()` may not be the best way to get your result. (note that it is fine for all assignments for this class).  We can instead use a combination of `dbSendQuery()` and `fetch()` in a `while` loop to fetch results a thousand rows at a time. For more information, please refer to the R-Bootcamp, module 5.

## Running a Simple Query

Let's ask a simple question about the `patient` table. How many female patients are there? There are two ways we can do this. The first would be to select those patients that are female with a `SELECT` statement, return those patients as a `data.frame`, and then just count the number of rows in the `data.frame`.

```{r}
sqlStatement <- "SELECT * from patient WHERE Gender == 'Female'"

femalePatients <- dbGetQuery(SQLiteConnection, sqlStatement)

nrow(femalePatients)
```

Or, we can combine our statement with the `count` statement from above and get the answer directly.

```{r}
#note that for matching strings, we have to use single quotes in our SQL
#statement. Otherwise, R gets confused where the statement begins or ends.
sqlStatement <- "SELECT COUNT(*) FROM patient WHERE Gender == 'Female'"

numberFemalePatients <- dbGetQuery(SQLiteConnection, sqlStatement)

numberFemalePatients
```

If we wanted the number of Female patients with a normal `riskCat`, we can chain these queries using `AND`.:

```{r}
#Again, enclosing strings within single quotes so we don't confuse R
sqlStatement <- "SELECT count(*) FROM patient WHERE GENDER == 'Female' 
                  AND riskCat == 'Normal'"

numberFemaleNormalPatients <- dbGetQuery(SQLiteConnection, sqlStatement)

numberFemaleNormalPatients
```

If we only wanted a few columns from the table (such as `patientid`, `GENDER` and `riskCat`), we can change our select statement to only return these columns:

```{r}
sqlStatement <- "SELECT patientid, GENDER, riskCat FROM patient 
            WHERE GENDER == 'Female' AND riskCat == 'Normal'"

selectResults <- dbGetQuery(SQLiteConnection, sqlStatement)

selectResults[1:10,]
```

##Assignment Week 3 (2 Points)

Load the rest of your tables into your SQLite Database. Remember, you will need to disconnect from the database using `dbDisconnect()` to save the tables into the database. You can do this after you write all the tables into the database. 

Construct a query for each table to confirm that the number of rows of each loaded table is identical to the number of rows in the corresponding table you loaded into R. 

Submit all of your loading and querying code using the `week3Submission.Rmd` file. 

##For More Information

Please refer to [R-Bootcamp Module 5](https://github.com/laderast/r-bootcamp/tree/master/module5) for more information on databases and SQLite.
